Evaluating Perceptual Judgements on 3D Printed Bar Charts

### 2024-01-22 jds.tex
point-by-point response

Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion\citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in \citet{fig:sine-illusion}).
While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception. Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.

@@

In this paper, we discuss a study designed to examine Cleveland \& McGill's experiments on grouped bar charts using modern graphics in two and three dimensions. This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data. In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information.
In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from \citet{clevelandGraphics1984} in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs. The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.

@@

\begin{figure}
\includegraphics[width=0.8\linewidth]{jds_files/figure-latex/midmeans-log-errors-1} \caption{Midmeans and observed values of log absolute errors for the true ratio of bars. Summary lines are computed from raw data using a loess smooth.}\label{fig:midmeans-log-errors}
\end{figure}

@@

\hypertarget{linear-mixed-effects-model}{%
\subsection{Linear Mixed Effects Model}\label{linear-mixed-effects-model}}

In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a linear mixed effects model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio. This allows us to test for significant differences (though given the midmean log absolute error plots, we do not expect to find any) as well as to quantify effect sizes for future studies. The formal statistical model is as follows:

\[y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}\]

@@

\begin{itemize}
\item
  \(y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\)
\item
  \(S_i\sim N(0,\sigma^2_S)\) is the random effect of the \(i^{th}\) subject
\item
  \(R_j\) is the effect of the \(j^{th}\) ratio
\item
  \(G_k\) is the effect of the \(k^{th}\) display type
\item
  \(T_l\) is the effect of the \(l^{th}\) comparison type
\item
  \(\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)\) is the random error
\end{itemize}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:unnamed-chunk-2}Analysis of Fixed Effects}

\begin{tabular}{llll}
\toprule
Term & \multicolumn{1}{c}{$\hat{\beta}$} & \multicolumn{1}{c}{95\% CI} & \multicolumn{1}{c}{$t$}\\
\midrule
Intercept & 2.49 & {}[1.94, 3.04] & 8.84\\
Ratio & 0.02 & {}[-0.04, 0.07] & 0.56\\
Plot3dDigital & 0.15 & {}[-0.14, 0.43] & 1.02\\
Plot3dPrint & -0.06 & {}[-0.35, 0.22] & -0.44\\
Type & 0.11 & {}[-0.13, 0.35] & 0.94\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

No differences were detected for the true ratio of bars (\(t = 0.5590574\)), whether the bars were adjacent or separated (\(t = 0.9381629\)), or for the plot nested within the true ratio (\(F(2, 441) = 1.1110313\)).
While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes.

>>>>

Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion\citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in \Cref{fig:sine-illusion}.
While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception. Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.

@@

In this paper, we discuss a study designed to examine Cleveland \& McGill's experiments on grouped bar charts using modern graphics in two and three dimensions. This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data. In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information.
In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from \citet{clevelandGraphical1984} in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs. The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.

@@

\begin{figure}
\includegraphics[width=\textwidth]{jds_files/figure-latex/midmeans-log-errors-1} \caption{Midmeans and observed values of log absolute errors for the true ratio of bars. Summary lines are computed from raw data using a loess smooth.}\label{fig:midmeans-log-errors}
\end{figure}

@@

\hypertarget{generalized-additive-model}{%
\subsection{Generalized Additive Model}\label{generalized-additive-model}}

In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio. The formal statistical model is as follows:

\[y_{ijklm}= s(R_j * T_l) + s(R_j *G_k) + T_l + G_k + s(S_i) + \epsilon_{ijklm}\]

@@

\begin{itemize}
\item
  \(y_{ijklm}=\log_2(|\text{Judged Percent} - \text{True Percent}|+1/8)\)
\item
  \(s(S_i)\) is the random effect of the \(i^{th}\) subject
\item
  \(s(R_j*T_l)\) is the effect of the \(j^{th}\) ratio, by comparison type
\item
  \(s(R_j*G_k)\) is the effect of the \(j^{th}\) ratio, by display type
\item
  \(T_l\) is the effect of the \(l^{th}\) comparison type
\item
  \(G_k\) is the effect of the \(k^{th}\) display type
\item
  \(\epsilon_{ijklm}\sim N(0,\sigma^2_\epsilon)\) is the random error
\end{itemize}

A generalized additive model with random effects for subject was fit using the \texttt{mgcv} package \citep{mgcv1, mgcv2, mgcv3}. Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}. It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in \citet{hollandsBiasProportionJudgments2000}.

While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 trials (8.79\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562). This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5. In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types. It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum. A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.

\begin{table}

\caption{\label{tab:gam-param}Parametric coefficients in gam model.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
  & Estimate & Std. Error & t value & Pr(>|t|)\\
\midrule
(Intercept) & 2.6862 & 0.1780 & 15.0931 & 0.0000\\
plot3dDigital & 0.1357 & 0.1409 & 0.9626 & 0.3363\\
plot3dPrint & -0.0563 & 0.1415 & -0.3980 & 0.6908\\
typeAdjacent & 0.0968 & 0.1197 & 0.8088 & 0.4191\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:gam-smooth}Approximate significance of smooth terms in gam model.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Smooth & edf & Ref.df & F & p-value\\
\midrule
s(Ratio):plot2dDigital & 0.0005 & 0.0009 & 0.0000 & 0.5000\\
s(Ratio):plot3dDigital & 1.7934 & 2.2056 & 1.2807 & 0.2481\\
s(Ratio):plot3dPrint & 1.0003 & 1.0005 & 2.4488 & 0.1183\\
s(Ratio):typeSeparated & 4.2371 & 4.7338 & 5.8074 & 0.0001\\
s(Ratio):typeAdjacent & 1.0007 & 1.0014 & 3.5452 & 0.0604\\
\addlinespace
s(participant) & 27.0449 & 34.0000 & 4.7264 & 0.0000\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}
\includegraphics[width=\textwidth]{jds_files/figure-latex/gam-pred-chart-1} \caption{Predictions with standard errors for fixed effects in a generalized additive model with splines for ratio by comparison and ratio by display method. It is clear that the separated comparisons are easier to estimate when small, right around 50\%, or large, and harder to estimate between these points. What is interesting is that no such trend is present for adjacent comparisons. }\label{fig:gam-pred-chart}
\end{figure}

While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g.~slider vs.~numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
